How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?

### 2021-09-27 analysis/paper/paper.Rmd

---
title: "How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?"
author:
  - Gayoung Park
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2: 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  We use models from human behavioral ecology to examine stone artifacts from 22 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene when new tools, such as stemmed points and blades, appeared in the archaeological record. We focus on two questions: what changes in foragers’ landscape use were associated with the introduction of new tools? And what changes in the way people use habitation sites and mobility were associated with the new technology? To answer these questions we present quantitative analyses of artifact volumetric density, retouch frequency, composition of toolkits, and artifact raw materials. We explore the environmental and demographic contexts by applying paleoclimate simulations and summed probability distribution models. We find that quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene. Our results show that forager groups using stemmed points were associated with occupation of marginal or extreme environments. In addition, groups with stemmed points were associated with expedient technologies, indicating residential and less mobile behaviours. The environmental context of this technological innovation was a gradual decrease in temperature into the LGM. A population increase followed after the appearance of stemmed points.

@@

The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (ca. 42-35 ka) in Korea is often assumed to have transformed forager lifeways because it was thought to have indicated more specialized hunting practices [@seong2008tanged; @chang2013human]. The stemmed point is a projectile point made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. These types of artifacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent a hafting projection used for joining the lithic to a wooden shaft. If these artifacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea. Most open sites in Korea do not preserve organic remains due to acidic soil. There are a few limestone cave sites containing human fossils or faunal remains (e.g. Durubong and Jeommal Cave), which located in constrained area to access in North Korea or have been rarely analyzed to directly apply to other archaeological research[@bae2013paleolithic]. Therefore stone artifacts are our only window to examine hunting and subsistence strategies. Here we use a behavioral ecological framework to explore questions about the implications of the appearance of this new technology. We used data from 22 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the of the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.

@@

Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin; @bae2017late]. To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail. Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artifact assemblages using Human Behavioral Ecology (HBE) concepts. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility; @seong2015diversity; @seong2008tanged], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that were part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.

@@

What makes the Korean Late Pleistocene technological transition distinctive from what we see at Shuidonggou and in Japan? The key difference is that the earliest signs of new technologies in Korean assemblages are stemmed points, followed by blade technologies. Stemmed points appeared in northern China and Japan after they appear in the Korean Peninsula. The oldest stemmed points so far in Korea are from the Bonggok site dated to ca. 41.5 ka, and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin]. After blades became widespread in Korea at ca. 27 ka, and were used to make the stemmed points, the shape of points became more standardized with one or two ridges on the dorsal side and triangular cross section. Replacing flakes with blades as blanks of the stemmed points led to an increase in quantity of the points over time. In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@bae2010origin; @lee2013current; @Lee_2016].

@@

Second, we ask what changes in the way people used habitation sites and mobility were associated with the new technology? After the appearance of these new tools, did people tend to be more or less mobile? The stone artifact assemblage from a site can inform us how long individuals or groups stayed and give insights into their activities [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface]. For example, stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) can be used as proxies to represent occupation patterns [@barton2011modeling; @clark2017lithics]. Assemblages with low density but a high proportion of retouched and backed pieces may indicate the remains of 'short-term camp', which is a site for small and ephemeral overnight camp or limited activity station. On the other hand, the combination of high density and low retouched pieces might be expected at logistically organized basecamp, which is a site with greater residential stability, long site occupation and occupied by larger groups [@clark2017lithics; @clark2019landscapes].

@@

After the first Paleolithic sites such as Dongkwanjin and Yondaebong were discovered during the mid-1930s, more than 200 Paleolithic sites have been discovered in South Korea [@norton2000current; @lee2017korean; @lee2019were]. We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change. This resulted in a sample of 33 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts. For example, the Hwadaeri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.

@@

The 33 assemblages consist of (non-blade) cores, blade cores, blades, flakes, debris, hammers, choppers, planes, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc. We excluded some artifacts from our sample because of uncertainty about their typology. The stone artifacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports. We have visited most of these collections to obtain permission to study the materials directly.

@@

In addition to these data on human behaviour from the stone artifact assemblages, our HBE models require environmental proxies to build hypotheses of behavioural change influenced by the surrounding environment. However, there are few paleoenvironmental proxies in Korea relevant to our study period. For example, geoarchaeological observations simply suggest that environment of the period was overall cooler and drier during MIS 3 [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late]. To supplement the limited availability of high-resolution local proxies, we used climate information from a simulated data set of paleoclimate including global monthly temperature, covering the last 120,000 years [@beyer2020high]. We extracted annual temperature values from the data generated by @beyer2020high for all site locations in our sample. In addition, we extracted site elevation values from a raster, as a proxy to represent different environments at local scales, to identify how these impact mobility and site occupation, and their relationship with the stemmed points.

@@

We used radiocarbon dates as a proxy for human population to infer the demographic context of technological change. Summed probability distributions (SPD) of radiocarbon dates are increasingly used to infer temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread]. Despite concerns about limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon; @carleton2021evaluating], several studies have shown good agreement between SPDs and other archaeological indicators (e.g. site counts, settlement size) so this method may allow a first approximation of population fluctuations [@french2015upper; @palmisano2017comparing]. Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes. For instance, @timpson2014reconstructing apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n = 243). We used `r nrow(dates_clean)` radiocarbon ages from `r nrow(korean_archaeological_site_locations)` sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.

@@

```{r volume-density, fig.cap="Artifact volumetric density over time. The size of data points represent the total number of artifacts from each assemblage and the colour indicates the presence of stemmed points. The inset plot shows direct comparison of artifact densities for assemblages with and without stemmed points.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/002-artifact-volumetric-density.R"))}
quietly(source(here::here("analysis/paper/002-artifact-volumetric-density.R")))
knitr::include_graphics(here::here("analysis/figures/002-age-by-density.png"))
```

@@

In \@ref(fig:artifact-composition) we see that side scrapers are the major part of most assemblages and are present in all assemblages. For example, more than 70 percent of the two assemblages from Hwaderi are side scrapers. The proportion of side scrapers shows a striking trend over time, rising to a peak at 38-39 ka, and declining thereafter. In addition to side scrapers, cores also exist in all assemblages, often as a high proportion. The proportion of choppers is lower in more recent assemblages. Stemmed points first appeared in Bonggok around 41.5 ka and were typically found in assemblages that contain blades, except for Songamri-A. The earliest blades are also recorded from Bonggok but it is not guaranteed that those artifacts as well as other earlier blades are actual blades or elongate flakes without accompanying blade core. Blade cores first appeared in Suyanggae-C at 34.3 ka. Other artifacts, including burins, denticulates, end scrapers, handaxes, knives, notches, planes, and polyhedrals made up a small proportion of the assemblages and show no directional chronological change.

```{r artifact-composition, fig.cap="The composition of each assemblage. We excluded artifacts related to manufacturing processes such as pebbles, hammers, flakes and debris, and artifacts that appeared only in a few assemblages including points, beak shaped pieces, and awls, and unknown pieces. The colour represents different types of tools and the assemblages are placed in chronological order.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/004-artifact-composition.R"))}
quietly(source(here::here("analysis/paper/004-artifact-composition.R")))
knitr::include_graphics(here::here("analysis/figures/004-artifact-types.png"))
```

@@

Quartz, quartz vein and quartzites were the most frequently found raw materials, and they were constantly used throughout the Late Pleistocene (Figure \@ref(fig:raw-materials)). For example, both Galdamri-A, dated to 44 ka with a small assemblage, and Hopyeongdong-B, dated to 27.5 ka with a larger assemblage, consist of only the quartz related materials. The use of chert, hornfels, rhyolite, and shale increased after 41 ka, but these raw materials remained a small proportion of assemblages until 25 ka. For example, hornfels are dominant at Yongsandong and Yongbang, dated to 25 ka. Chert, hornfels, rhyolite, and shale are very suitable to make elongate blades and stemmed points because of the predictability of flaking afforded by their fine-grained texture. For example, Yongsandong has the largest number of stemmed points (n=40) among the assemblages studied here, and all stemmed points from Yongsandong are made from hornfels. Other raw materials, such as porphyry and sandstone, were found in just a few assemblages in small proportions.

```{r raw-materials, fig.cap="The composition of raw material type in each assemblage. We excluded raw materials that are included in less than 5 assemblages including crystal, basalt, iron_ore, slate, limestone, granite, gneiss, tuff, and unidentified ones. Hornfels are shown in 4 assemblages but since it occupies more than half of the Yongsandong and Yongband assemblages, we included it in the plot. The color represents different types of raw materials, and the assemblages are placed in chronological order.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/005-raw-materials.R"))}
quietly(source(here::here("analysis/paper/005-raw-materials.R")))
knitr::include_graphics(here::here("analysis/figures/005-raw-materials.png"))
```

@@

Figure \@ref(fig:site-elevation) shows that people occupied sites located at similar elevations, around 0 to 200 meters above sea level. As we saw for artifact density, there is an increase in the upper bound of site elevation in more recent times. The most striking pattern in the elevation data is the substantial difference in elevation between sites with stemmed points and sites without. The distribution of elevations of sites without stemmed points is generally lower than the elevations of sites with stemmed points. Although not a statistically significant difference in elevations (`r elev_sp_ttest_str`), this may indicate that forager groups who used stemmed points were generally more able to occupy higher altitudes, while the groups without stemmed points tended to prefer lower altitudes. According to our locally weighted regression, sites without stemmed points reach a maximum elevation of about 150 m at around 39 ka. Sites containing stemmed points reach a maximum elevation of 175 m at 28 ka.

@@

Figure \@ref(fig:spd-model) shows the summed probability distribution (SPD) of the radiocarbon ages from the sites in our sample overlaid with distributions (indicated by the grey envelopes) generated from four models of demographic change for the period 50 ka to 10 ka (exponential, uniform, linear, and logistic). The fit of models to our observed SPD was computed using 1000 simulations, and the resulting global p-values indicate significant deviations in the observed SPD from the modeled distributions. Overall, the linear model is the best fit with the observed SPD, having the lowest AIC value of the three models where we could compute this. We can say that the population did not grow or decline strongly logistically, exponentially or linearly during the Late Pleistocene, but instead followed a noisy distribution.  The most striking results here are the SPD showing a steady growth rising to a peak at about 35 ka, indicated by the significant deviations from all four null models (shaded in red). The SPD remains high with minor deviations until declining to a lower level at 32 ka. After 32 ka the SPD remains generally low, with frequent temporary fluctuations.

```{r spd-model, fig.cap=paste0("Summed probability distribution of ", nrow(dates_clean), " radiocarbon dates. The black solid line represents actual radiocarbon ages. The grey shaded region shows the Monte Carlo envelope that encompasses the 95% confidence interval for the null models. The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected. We applied logistic, exponential and linear to AIC along with the logistical growth model (fit) to evaluate each model [@riris2019widespread]. We excluded the uniform model because it has no fitted curve or surface to evaluate. The best fit model has the lowest AIC score. In this case, the linear model is the best fit for the observed data"), cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/009-summed-probability-distributions.R"))}
# quietly(source(here::here("analysis/paper/009-summed-probability-distributions.R")))
knitr::include_graphics(here::here("analysis/figures/008-summed-probability-distribution-models.png"))
```

@@

Our study focused on two questions: (1) what changes in foragers’ landscape use were associated with the introduction of the new tools?; and (2) what changes in people’s mobility and use of habitation sites were associated with the new technology? Overall, our results show that assemblages with stemmed points tend to have higher artifact densities and lower proportions of retouched pieces. Quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages throughout the late Pleistocene. The environmental context of this technological innovation was a gradual decrease in temperature into the LGM. And the population increase followed after the appearance of stemmed points.

@@

We examined the artifact volumetric density and retouch frequency to understand the difference in mobility strategies and site occupation patterns associated with the introduction of stemmed points. The analyses are based on the premise that highly mobile groups carried lightweight toolkits with more retouched and easily replaceable composite tools, and left smaller amounts of artifacts at sites as a result of short-term occupations. These curated technologies reflect an investment of time and effort in stone artifact manufacturing and maintenance to minimize travel costs of carrying the tools for high mobility or special purpose foray groups. On the other hand, less mobile groups produced fewer retouched tools and left higher densities of artifacts generated from long-term occupations. These expedient technologies are more typical for groups with greater residential stability who can make, use, abandon, and remake their tools frequently in the same location [@binford1979organization; @torrence1989time; @kuhn1994formal; @meignen2006middle; @kuhn2014mousterian; @vaquero2018searching; @clark2019landscapes]. Our data support the general concept of the curated and expedient technology by showing a strong pattern of artifacts from less dense assemblages having higher retouch frequencies, while more dense assemblages have lower retouch frequencies. Our results show that foragers with stemmed points tended to stay in the same site for longer periods with higher artifact densities and fewer retouched pieces, compared to assemblages without stemmed points. In general, the new technology of stemmed points was more associated with expedient technological strategies.

>>>>

---
title: "How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?"
author:
  - Gayoung Park
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2: 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  We use models from human behavioral ecology to examine stone artifacts from 23 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene when new tools, such as stemmed points and blades, appeared in the archaeological record. We focus on two questions: what changes in foragers’ landscape use were associated with the introduction of new tools? And what changes in the way people use habitation sites and mobility were associated with the new technology? To answer these questions we present quantitative analyses of artifact volumetric density, retouch frequency, composition of toolkits, and artifact raw materials. We explore the environmental and demographic contexts by applying paleoclimate simulations and summed probability distribution models. We find that quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene. Our results show that forager groups using stemmed points were associated with occupation of marginal or extreme environments. In addition, groups with stemmed points were associated with expedient technologies, indicating residential and less mobile behaviours. The environmental context of this technological innovation was a gradual decrease in temperature into the LGM. A population steadily increased before the appearance of stemmed points.

@@

The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (ca. 42-35 ka) in Korea is often assumed to have transformed forager lifeways because it was thought to have indicated more specialized hunting practices [@seong2008tanged; @chang2013human]. The stemmed point is a projectile point made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. These types of artifacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent a hafting projection used for joining the lithic to a wooden shaft. If these artifacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea. Most open sites in Korea do not preserve organic remains due to acidic soil. There are a few limestone cave sites containing human fossils or faunal remains (e.g. Durubong and Jeommal Cave), which located in constrained area to access in North Korea or have been rarely analyzed to directly apply to other archaeological research[@bae2013paleolithic]. Therefore stone artifacts are our only window to examine hunting and subsistence strategies. Here we use a behavioral ecological framework to explore questions about the implications of the appearance of this new technology. We used data from 23 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the of the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.

@@

Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin; @seong2016eastern; @bae2017late]. To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail. Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artifact assemblages using Human Behavioral Ecology (HBE) concepts. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility; @seong2015diversity; @seong2008tanged], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that were part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.

@@

What makes the Korean Late Pleistocene technological transition distinctive from what we see at Shuidonggou and in Japan? The key difference is that the earliest signs of new technologies in Korean assemblages are stemmed points, followed by blade technologies. Stemmed points appeared in northern China and Japan after they appear in the Korean Peninsula. The stemmed points from the Bonggok site had been known as the oldest stemmed points dated to ca. 41.5 ka, and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin]. After blades became widespread in Korea at ca. 27 ka, and were used to make the stemmed points, the shape of points became more standardized with one or two ridges on the dorsal side and triangular cross section. Replacing flakes with blades as blanks of the stemmed points led to an increase in quantity of the points over time. In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@bae2010origin; @lee2013current; @Lee_2016]. However, recently excavated Hajinri site contain stemmed points and blades dated to 42 ka, which is even older than Bonggok site.

@@

Second, we ask what changes in the way people used habitation sites and mobility were associated with the new technology? After the appearance of these new tools, did people tend to be more or less mobile? The stone artifact assemblage from a site can inform us how long individuals or groups stayed and give insights into their activities [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface]. For example, stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) can be used as proxies to represent occupation patterns [@barton2011modeling; @clark2017lithics]. Assemblages with low density but a high proportion of retouched and backed pieces may indicate the remains of 'short-term camp', which is a site for small and ephemeral overnight camp or limited activity station. On the other hand, the combination of high density and low retouched pieces might be expected at logistically organized 'base camp', which is a site with greater residential stability, long site occupation and occupied by larger groups [@clark2017lithics; @clark2019landscapes].

@@

After the first Paleolithic sites such as Dongkwanjin and Yondaebong were discovered during the mid-1930s, more than 200 Paleolithic sites have been discovered in South Korea [@norton2000current; @lee2017korean; @lee2019were]. We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change. We acknowledge the limitations of directly accepting the dating results from excavation reports since the majority of sites in Korea are open-air localities which are more prone to be disturbed compared to cave sites. Therefore, we carefully excluded sites whose assemblages are not matched with their dating results. 
This resulted in a sample of 35 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts. For example, the Hwadaeri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.

@@

The 35 assemblages consist of (non-blade) cores, blade cores, blades, flakes, debris, hammers, choppers, planes, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc. We excluded some artifacts from our sample because of uncertainty about their typology. The stone artifacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports. We have visited most of these collections to obtain permission to study the materials directly.

@@

In addition to these data on human behavior from the stone artifact assemblages, our HBE models require environmental proxies to build hypotheses of behavioral change influenced by the surrounding environment. However, there are few paleoenvironmental proxies in Korea relevant to our study period. For example, geoarchaeological observations simply suggest that environment of the period was overall cooler and drier during MIS 3 [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late]. To supplement the limited availability of high-resolution local proxies, we used climate information from a simulated data set of paleoclimate including global monthly temperature, covering the last 120,000 years [@beyer2020high]. We extracted annual temperature values from the data generated by @beyer2020high for all site locations in our sample. In addition, we extracted site elevation values from a raster, as a proxy to represent different environments at local scales, to identify how these impact mobility and site occupation, and their relationship with the stemmed points.

@@

We used radiocarbon dates as a proxy for human population to infer the demographic context of technological change. Summed probability distributions (SPD) of radiocarbon dates are increasingly used to infer temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread]. Despite concerns about limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon; @carleton2021evaluating], several studies have shown good agreement between SPDs and other archaeological indicators (e.g. site counts, settlement size) so this method may allow a first approximation of population fluctuations [@french2015upper; @palmisano2017comparing]. Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes. For instance, @timpson2014reconstructing apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n = 243). In addition to the 93 radiocarbon ages from our targeted sites, we included all available dating results (540 ages) from other Korean Paleolithic sites that have younger and older assemblages. We compiled a total of 633 dates from 113 sites and applied a chronometric hygiene protocol to assess the reliability of radiocarbon determinations in relation to target events. Following the criteria from @napolitano2019reevaluating and @fitzpatrick2006critical, we kept ages from dating short-lived terrestrial material (i.e., plant remains) and charcoal, and having laboratory name and number. We excluded OSL dates and sites with only one date. We ended up using `r nrow(dates_clean)` radiocarbon ages from `r nrow(korean_archaeological_site_locations)` sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.

@@

```{r volume-density, fig.cap="Artifact volumetric density over time. The size of data points represent the total number of artifacts from each assemblage and the colour indicates the presence of stemmed points. The inset plot shows direct comparison of artifact densities for assemblages with and without stemmed points.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/002-artifact-volumetric-density.R"))}
# quietly(source(here::here("analysis/paper/002-artifact-volumetric-density.R")))
knitr::include_graphics(here::here("analysis/figures/002-age-by-density.png"))
```

@@

In \@ref(fig:artifact-composition) we see that side scrapers are the major part of most assemblages and are present in all assemblages. For example, more than 70 percent of the two assemblages from Hwaderi are side scrapers. The proportion of side scrapers shows a striking trend over time, rising to a peak at 38-39 ka, and declining thereafter. In addition to side scrapers, cores also exist in all assemblages, often as a high proportion. The proportion of choppers is lower in more recent assemblages.
Stemmed points first appeared in Hajinri around 42 ka and were typically found in assemblages that contain blades, except for Songamri-A. The earliest blades and blade cores are also recorded from Hajinri but the date of this site can be unreliable because the stemmed points and blades are highly standardized and refined. Bonggok site, which had been known as the earliest site for stemmed points, also includes blades but it has not been guaranteed that those artifacts as well as other earlier blades are actual blades or elongate flakes without accompanying blade core. Except for Hajinri, blade cores first appeared in Suyanggae-C at 34.3 ka. Other artifacts, including burins, denticulates, end scrapers, handaxes, knives, notches, planes, and polyhedrals made up a small proportion of the assemblages and show no directional chronological change.

```{r artifact-composition, fig.cap="The composition of each assemblage. We excluded artifacts related to manufacturing processes such as pebbles, hammers, flakes and debris, and artifacts that appeared only in a few assemblages including points, beak shaped pieces, awls, and anvils, and unknown pieces. The colour represents different types of tools and the assemblages are placed in chronological order.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/004-artifact-composition.R"))}
quietly(source(here::here("analysis/paper/004-artifact-composition.R")))
knitr::include_graphics(here::here("analysis/figures/004-artifact-types.png"))
```

@@

Quartz, quartz vein and quartzites were the most frequently found raw materials, and they were constantly used throughout the Late Pleistocene (Figure \@ref(fig:raw-materials)). For example, both Galdamri-A, dated to 44 ka with a small assemblage, and Hopyeongdong-B, dated to 27.5 ka with a larger assemblage, consist of only the quartz related materials. The use of chert, hornfels, rhyolite, and shale increased after 42 ka, but these raw materials remained a small proportion of assemblages until 25 ka. They are very suitable to make elongate blades and stemmed points because of the predictability of flaking afforded by their fine-grained texture. For example, Hajinri has the largest number of stemmed points (n=86) among the assemblages studied here, and most stemmed points from Hajinri are made from shales. Other raw materials, such as porphyry and sandstone, were found in just a few assemblages in small proportions.

```{r raw-materials, fig.cap="The composition of raw material type in each assemblage. We excluded raw materials that are included in less than 5 assemblages including crystal, basalt, iron_ore, slate, limestone, granite, gneiss, tuff, amphibolite, andesite, grit, metamorphic-rock, quartzitic-gneiss, siliceous-schist, and unidentified ones. Hornfels are shown in 4 assemblages but since it occupies more than half of the Yongsandong and Yongband assemblages, we included it in the plot. The color represents different types of raw materials, and the assemblages are placed in chronological order.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/005-raw-materials.R"))}
quietly(source(here::here("analysis/paper/005-raw-materials.R")))
knitr::include_graphics(here::here("analysis/figures/005-raw-materials.png"))
```

@@

Figure \@ref(fig:site-elevation) shows that people occupied sites located at similar elevations, around 0 to 200 meters above sea level. As we saw for artifact density, there is an increase in the upper bound of site elevation in more recent times. The most striking pattern in the elevation data is the substantial difference in elevation between sites with stemmed points and sites without. The distribution of elevations of sites without stemmed points is generally lower than the elevations of sites with stemmed points. Although not a statistically significant difference in elevations (`r elev_sp_ttest_str`), this may indicate that forager groups who used stemmed points were generally more able to occupy higher altitudes, while the groups without stemmed points tended to prefer lower altitudes. According to our locally weighted regression, sites without stemmed points reach a maximum elevation of about 150 m at around 39 ka. Sites containing stemmed points reach a maximum elevation of 175 m at 28 ka. We acknowledged that site elevation might not be the best proxy for the environmental context due to mountainous geography of Korea and possible disturbance during the site formation processes. However recent studies show that elevation makes huge impact on the diversity and ecology of plants and animals in Korea [@sohn2019effect; @park2020temporal] and thus enable us to explore the elevation as a environmental proxy.

@@

Figure \@ref(fig:spd-model) shows the summed probability distribution (SPD) of the radiocarbon ages from the sites in our sample overlaid with distributions (indicated by the grey envelopes) generated from four models of demographic change for the period 50 ka to 10 ka (linear, exponential, logistic, and uniform). The fit of models to our observed SPD was computed using 1000 simulations, and the resulting global p-values indicate significant deviations in the observed SPD from the modeled distributions. Overall, the linear model is the best fit with the observed SPD, having the lowest AIC value of the three models where we could compute this. We can say that the population did not grow or decline strongly logistically, exponentially or linearly during the Late Pleistocene, but instead followed a noisy distribution. The most striking results here are the SPD showing a steady growth rising to a peak at about 43 ka, indicated by the positive deviations from the three null models (shaded in red), except for the logistic model. During the transition period, SPD hits another peak at about 35ka. Then the SPD goes down with minor deviations to a lower level at 31.5 ka. From 27 ka the SPD constantly fluctuates with temporary positive deviation until 20ka.

```{r spd-model, fig.cap=paste0("Summed probability distribution of ", nrow(dates_clean), " radiocarbon dates. The black solid line represents actual radiocarbon ages. The grey shaded region shows the Monte Carlo envelope that encompasses the 95% confidence interval for the null models. The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected. We applied linear, exponential and logistic to AIC along with the logistical growth model (fit) to evaluate each model [@riris2019widespread]. We excluded the uniform model because it has no fitted curve or surface to evaluate. The best fit model has the lowest AIC score. In this case, the linear model is the best fit for the observed data"), cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/009-summed-probability-distributions.R"))}
# quietly(source(here::here("analysis/paper/009-summed-probability-distributions.R")))
knitr::include_graphics(here::here("analysis/figures/008-summed-probability-distribution-models.png"))
```

@@

Our study focused on two questions: (1) what changes in foragers’ landscape use were associated with the introduction of the new tools?; and (2) what changes in people’s mobility and use of habitation sites were associated with the new technology? Overall, our results show that assemblages with stemmed points tend to have higher artifact densities and lower proportions of retouched pieces. Quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages throughout the late Pleistocene. The environmental context of this technological innovation was a gradual decrease in temperature into the LGM. And the population increase before the appearance of stemmed points.

@@

We examined the artifact volumetric density and retouch frequency to understand the difference in mobility strategies and site occupation patterns associated with the introduction of stemmed points. The analyses are based on the premise that highly mobile groups carried lightweight toolkits with more retouched and easily replaceable composite tools, and left smaller amounts of artifacts at sites as a result of short-term occupations. These curated technologies reflect an investment of time and effort in stone artifact manufacturing and maintenance to minimize travel costs of carrying the tools for high mobility or special purpose foray groups. On the other hand, less mobile groups produced fewer retouched tools and left higher densities of artifacts generated from long-term occupations. These expedient technologies are more typical for groups with greater residential stability who can make, use, abandon, and remake their tools frequently in the same location [@binford1979organization; @torrence1989time; @kuhn1994formal; @meignen2006middle; @kuhn2014mousterian; @vaquero2018searching; @clark2019landscapes]. As mentioned above, the sites we addressed here are all open-air localities with limited material remains to infer the clear view of subsistence strategies but our data still supports the general concept of the curated and expedient technology by showing a strong pattern of artifacts from less dense assemblages having higher retouch frequencies, while more dense assemblages have lower retouch frequencies. Our results show that foragers with stemmed points tended to stay in the same site for longer periods with higher artifact densities and fewer retouched pieces, compared to assemblages without stemmed points. In general, the new technology of stemmed points was more associated with expedient technological strategies.
